Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/overview.tex
related work

When applied to the staged code in \ref{fig:qs-staged}, our splitting
algorithm yields a two-pass program that uses the binary-search-tree
based implementation.  Specifically, in its first pass, the two-pass
program takes the input list and constructs a probabilistically
balanced binary search tree, which is isomorphic to a treap data
structure~\cite{treaps}.  In the second pass, the program performs,
for each rank, a binary search tree lookup, by walking the tree to
find the element with the desired rank.
%
\ref{fig:qs-staged} illustrates the code for the first and second
passes \texttt{qs1} and \texttt{qs2} of the staged function
\texttt{qss}.

>>>>

When applied to the staged code in \ref{fig:qs-staged}, our splitting
algorithm yields a two-pass program that uses the binary-search-tree
based implementation.  Specifically, in its first pass, the two-pass
program takes the input list and constructs a probabilistically
balanced binary search tree, which is isomorphic to a treap data
structure~\cite{SeidelAr96}.  In the second pass, the program performs,
for each rank, a binary search tree lookup, by walking the tree to
find the element with the desired rank.
%
\ref{fig:qs-staged} illustrates the code for the first and second
passes \texttt{qs1} and \texttt{qs2} of the staged function
\texttt{qss}.

### 2015-02-27 paper/relatedwork.tex
fixing various broken things
related work

\section{Related Work}

As Jorring and Scherlis argue frequency reduction and precomputation
are common techniques for both designing algorithms and performing
compiler optimizations~\cite{JS86-staging}.  The idea behind frequency
reduction is to identify computations that are performed multiple
times and pull them ahead so that they can be performed once and used
later as needed.  The idea behind precomputation is to identify
computations that can be performed earlier, for example at compile
time if their inputs are available statically and perform them at that
earlier time.  Many interesting algorithm design techniques and
compiler optimizations perform frequency reduction and precomputation.
For example, dynamic programming and loop hoisting are a frequency
reduction techniques, whereas dynamic data structures and incremental
computation are precomputation techniques.

Perhaps one of the most well studied example of precomputation is
partial evaluation.  Partial evaluation distinguishes between a static
(compile-time) and dynamic (run-time) stage. Given a program and the
values of its static inputs, which are the subset of the inputs that
become available at compilation time, partial evaluation generates a
specialized program by performing the computations that depend on the
statics inputs~\cite{jones96}.  The roots of partial evaluation go
back to Kleene, who in 1952 proved that a multivariate program
(function) $f(x_1, \ldots, x_{n+m})$ can be specialized for any given
$a_1 \ldots a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such
that $f(a_1, \ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots,
y_n)$~\cite{Kleene52}.  Kleene did not intend to improve efficiency,
however.  Lombardi appears to be the first person to use the term in
his work on incremental computation~\cite{Lombardi67}.  Since then,
there has a lot of work on partial evaluation, only broad-strokes of
which we can mention in this here.  We refer the interested reader to
the excellent book by Jones, Gommard, and Sestoft for a comprehesive
discussion of work until early 90's~\cite{JGS93}.

@@

The type-system component of our work is closely related to this later
line of work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose.  The key difference between our work and this prior work is
the staging or the splitting algorithm, which given a two-staged
program splits it into two stages that can be executed sequentially,
passing information from the earlier to the latter stage by
construction. 
%
\ur{Do prior work evaluate first-stage expressions inside second stage
expressions?}
%
Unlike in partial evaluation, our splitting algorithm requires no
knowledge of the arguments themselves. The only information that is
required is their staging. Using just the staging information, the
splitting algorithm can generate programs that can perform non-trivial
transformations on the code automatically.  As our examples illustrate
such staging transformations can improve the efficiency of the source
program, sometimes in asymptotically significant ways.  
We are not aware of prior work in the context of higher-order
functional languages that presents such a splitting algorithm.

Perhaps the closest work to ours is algorithm for ``data
specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  As our
work, Knoblock and Ruf are concerned about the problem of splitting an
explicitly staged program into two stages.  However, they only
consider, simple first-order language and straight-line
programs. Automatic pass seperation has also been a topic of interest
in programming systems for computer graphics, where, to achieve
high-performance, programs are typically manually separated into
components by frequency of execution (Seperation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass seperation to the compiler. However, all
prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
limited to simple, imperative languages.

\paragraph{Memoization.} Leaves code structure unchanged.

\paragraph{Incremental computation.} Usually not applicable within a
computation, require ``mutator'' ``core'' behavior.

>>>>

\section{Related Work}
\label{sec:related}

As Jorring and Scherlis argue frequency reduction and precomputation
are common techniques for both designing algorithms and performing
compiler optimizations~\cite{JS86-staging}.  The idea behind frequency
reduction is to identify computations that are performed multiple
times and pull them ahead so that they can be performed once and used
later as needed.  For example, dynamic programming and loop hoisting,
as well as the work presented here are a frequency reduction
techniques.
%
The idea behind precomputation is to identify computations that can be
performed earlier, for example at compile time if their inputs are
available statically and perform them at that earlier time.  For
example, dynamic algorithms, partial evaluation, and incremental
computation are all precomputation techniques.

\paragraph{Precomputation techniques.}
Perhaps one of the most well studied example of precomputation is
partial evaluation.  Partial evaluation distinguishes between a static
(compile-time) and dynamic (run-time) stage. Given a program and the
values of its static inputs, which are the subset of the inputs that
become available at compilation time, partial evaluation generates a
specialized program by performing the computations that depend on the
statics inputs~\cite{jones96}.  The roots of partial evaluation go
back to Kleene, who in 1952 proved that a multivariate program
(function) $f(x_1, \ldots, x_{n+m})$ can be specialized for any given
$a_1 \ldots a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such
that $f(a_1, \ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots,
y_n)$~\cite{Kleene52}.  Kleene did not intend to improve efficiency,
however.  Lombardi appears to be the first person to use the term in
his work on incremental computation~\cite{Lombardi67}.  Since then,
there has a lot of work on partial evaluation, only broad-strokes of
which we can mention in this here.  We refer the interested reader to
the excellent book by Jones, Gommard, and Sestoft for a comprehesive
discussion of work until early 90's~\cite{JGS93}.

@@

The type-system component of our work is closely related to this later
line of work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose.  The key difference between our work and this prior work is
the staging or the splitting algorithm, which given a two-staged
program splits it into two stages that can be executed sequentially,
passing information from the earlier to the latter stage by
construction. 
%
Unlike in partial evaluation, our splitting algorithm requires no
knowledge of the arguments themselves. The only information that is
required is their staging. Using just the staging information, the
splitting algorithm can generate programs that can perform non-trivial
transformations on the code automatically.  As our examples illustrate
such staging transformations can improve the efficiency of the source
program, sometimes in asymptotically significant ways.  
We are not aware of prior work in the context of higher-order
functional languages that presents such a splitting algorithm.

An important class of pre-computation techniques include incremental
computation, where a program can respond to small changes to its input
efficiently by recomputing only the parts of the computation affected
by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
As other pre-computation techniques, incremental computation can be
thought as a staged model, where changes (inputs) may come over a
period of time in stages. 
%
It does not, however, fix any of the inputs or anticipate the future
inputs, allowing in the most general case for all data to change. As a
result, its benefits are dependent on the data instance (nature of the
inputs). In addition, incremental computation techniques typically
utilize significant run-time (dynamic) information, because no
assumptions about input instances are made.  For example, we can apply
incremental computation to our quickselect example
(ref{sec:overview}), but we would not be able to achieve the bounds
that we did because incremental computation techniques would unfold
the quickselect function based on the demanded ranks, which can
require linear time (as opposed to logarithmic) at each stage.  In
addition, such an implementation would maintain rather sophisticated
data structures dynamically at run time, the benefits of which depend
highly on the forms of changes applied.

\paragraph{Frequency-reduction techniques.}
Perhaps the closest work to ours is algorithm for ``data
specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  As our
work, Knoblock and Ruf are concerned about the problem of splitting an
explicitly staged program into two stages.  However, they only
consider, simple first-order language and straight-line
programs. Automatic pass seperation has also been a topic of interest
in programming systems for computer graphics, where, to achieve
high-performance, programs are typically manually separated into
components by frequency of execution (Seperation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass seperation to the compiler. However, all
prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
limited to simple, imperative languages.
